Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019

### 2020-05-07 manuscript/EP1162_manuscript_draft.Rmd
manny work

#### Volume estimates
Tree volumes were estimated using species- and region-specific equations [@nigh2016] based on tree height and DBH.  We estimated tree heights by fitting diameter-height models developed separately for fir and spruce, and for each period.  Trees with broken tops were omitted from diameter-height models.

@@

Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (p<0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Bl","Sx")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).  Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`appNums("htModelFit",display="cite")`).  The model slightly overestimated tree heights in shorter trees, and underestimated tree heights in taller trees.  This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@Nigh2016].

@@

# Create stand data
standDat<-
  dat %>% 
  group_by(Plot,Year) %>% 
  summarise(VolTot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
            VolMerch=sum(VolMerch,na.rm=T)*20,
            BA=sum(BA,na.rm=T)*20,
            k=n()*20) %>% 
  mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
  left_join(plotDat,by="Plot") %>%  # joinwith plot Data
   mutate(TU=paste("TU",BA.Target,sep=":"))  # create a TU column for plotting

@@

volPlot<-
  standDat %>% 
  # plot
  ggplot()+
  aes(x=Year,y=BA,fill=BA.Target)+
  geom_boxplot()+
  facet_wrap(~TU)+
  theme(legend.position="none")+
  ylab("Mean plot volume (m/hectare)")

volFit<-
  standDat %>% 
  lm(VolTot~BA.Target*Year,data=.)

@@

@@

`r figNums(name="sphSpeciesPlot",caption="Percentage change between 1992 and 2019 in stems per hectare (SPH) by diameter class, faceted by treatment unit and species.   tsphPCTPlotwo measurement periods.  Lowest value is included in each diameter class, highest value is not.")`

@@

@@

>>>>

#### Volume estimates
Tree volumes were estimated using species- and region-specific equations [@nigh2016] based on tree height and DBH.  We estimated tree heights by fitting a single diameter-height models developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.  Trees with broken tops were omitted from diameter-height models.

@@

Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (p<0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Bl","Sx")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).  Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`appNums("htModelFit",display="cite")`).  The model slightly overestimated tree heights in shorter trees, and underestimated tree heights in taller trees.  This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].

@@

# Create stand data
standDat<-
  dat %>% 
  group_by(Plot,Year) %>% 
  summarise(Vol.Tot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
            Vol.Merch=sum(VolMerch,na.rm=T)*20,
            BA=sum(BA,na.rm=T)*20,
            k=n()*20) %>% 
  mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
  left_join(plotDat,by="Plot") %>%  # joinwith plot Data
   mutate(TU=paste("TU",BA.Target,sep=":"))  # create a TU column for plotting

@@

volPlot<-
  standDat %>% 
  # plot
  ggplot()+
  aes(x=Year,y=BA,fill=BA.Target)+
  geom_boxplot()+
  facet_wrap(~TU)+
  theme(legend.position="none")+
  ylab("Plot volume (m³/hectare)")

# caption
figNums("volPlot.plot",display=FALSE)

volFit<-
  standDat %>% 
  lm(Vol.Tot~BA.Target*Year,data=.)

@@

### Stand increment
```{r function to compute stand increment}

standIncFunc<-function(varName) {

  standDat %>% 
  dplyr::select(Plot,Year,varName,BA.Target) %>% 
  pivot_wider(names_from="Year",values_from=varName) %>% 
  mutate(`1994Inc`=`1994`-`1992`) %>% # stand periodic increment 
  mutate(`1997Inc`=`1997`-`1994`) %>% # stand periodic increment 
  mutate(`2009Inc`=`2009`-`1997`) %>% # stand periodic increment 
  mutate(`2019Inc`=`2019`-`2009`) %>%  # stand periodic increment

  # divide by # growing seassons in each period
  mutate(`1994Inc`=`1994Inc`/3) %>% # 3 growing seasons in this period 
  mutate(`1997Inc`=`1997Inc`/3) %>% # 3 growing seasons in this period 
  mutate(`2009Inc`=`2009Inc`/11) %>% # 11 growing seasons in this period 
  mutate(`2019Inc`=`2019Inc`/11) %>% 
    dplyr::select(Plot,BA.Target,`1994`=`1994Inc`,`1997`=`1997Inc`,
                  `2009`=`2009Inc`,`2019`=`2019Inc`) %>% 
    pivot_longer(cols=`1994`:`2019`,names_to="Year",values_to="Inc") %>% 
    mutate(varname=varName) %>% 
    dplyr::select(Plot,BA.Target,varname,Year,Inc) %>% 
    return() # return
}

```{r apply stand increment function}

incPlot<-
  standIncFunc("Vol.Tot") %>%   # extract total volume 
  rbind(standIncFunc("Vol.Merch")) %>%   # extract total volume 
  rbind(standIncFunc("BA")) %>% # extract basal area
  drop_na(Inc) %>%

  # Create group summaries
  group_by(BA.Target,Year,varname) %>% 
  summarise(IncMean=mean(Inc),
            se=sd(Inc)/sqrt(length(Inc))) %>%

  # create lookup column
  mutate(TU=paste("TU",BA.Target,sep=":")) %>% # create lookup column

  # plotting
  ggplot()+
  aes(x=Year,y=IncMean,color=TU,group=1)+
  geom_errorbar(aes(ymin=IncMean-se,ymax=IncMean+se,width=.2))+
  geom_point(size=2)+
  geom_path()+
  facet_grid(varname~TU,scales="free_y")+
  theme(legend.position="none")+
  ylab("Annualized increment by period")

# caption
figNums("incPlot",display=FALSE)

Annual increment between the four periods differed between treatment units (`r figNums("incPlot",display="cite")`).  In the high-removal treatment unit, annual basal area increment increased in each subsequent measurement period, with the highest mean plot values of 0.9m^2^ per hectare.  In contrast, mean plot basal area annual increment was highest in the low-removal treatment unit during the 1994-1997 period, and decreased in subsequent periods.  A similar pattern was observed in the control treamtent unit.  Annual increment patterns in total and merchantable volumes were similar to basal area, with the exception of the high-removal treatment unit.  Here, maximum annual volume increment was highest between 1997 and 2009, with a mean plot value of 8m^3^ per year per hectare; this increment decreased slightly from 2009-2019 period.  In the control unit, volume increment from 2009-2019 was lower than from 1992-1994.

@@

`r figNums(name="sphPCTPlot",caption="Percentage change between 1992 and 2019 in stems per hectare (SPH) by diameter class, faceted by treatment unit and species.   tsphPCTPlotwo measurement periods.  Lowest value is included in each diameter class, highest value is not.")`

@@

```{r volume stand plots}
volPlot

`r figNums("volPlot.plot",caption="Volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`

```{r plot stand increment}

incPlot

`r figNums("incPlot",caption="Mean annual increment per plot over four measurement periods.  Top panel row shows basal area (BA), with units of m2 per hecatre.  Middle panel row shows merchantable volume (Vol.Merch), and bottom panel row shows total volume (Vol.Tot).  Volume units are m³ per hectare.")`

@@

# References
